---
title: "Exploratory analysis of ADL metabolomics"
output: "pdf_document"
author: "Aaron Yerke (aaronyerke@gmail.com)"
editor_options: 
  markdown: 
    wrap: 80
---

The purpose of this notebook is to explore the metabolomics from the study
below: BACKGROUND: Crossover studies can induce order effects, especially when
they lack a wash-22 out period.23 OBJECTIVE: We aimed to explore diet order
effects on energy balance and food intake24 between randomized diet order groups
in two inpatient crossover studies originally designed to25 compare within
subject differences in ad libitum energy intake between either minimally26
processed low carbohydrate (LC) versus low fat (LF) diets or macronutrient
matched diets27 composed of mostly minimally processed food (MPF) or
ultra-processed food (UPF).

Starting with the URINE 24h from the

```{r setup}
rm(list = ls()) #clear workspace
knitr::opts_knit$set(root.dir = normalizePath(".."))#Set root
knitr::opts_chunk$set(class.source = "Solarized Light")
```

```{r Load external libraries}
#### Loading libraries and data ####
#read in libraries
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("ggplot2", quietly = TRUE))  BiocManager::install("ggplot2")
library("ggplot2")
if (!requireNamespace("readxl", quietly = TRUE))  BiocManager::install("readxl")
library("readxl")
if (!requireNamespace("data.table", quietly = TRUE))  BiocManager::install("data.table")
library("data.table")
if(!require(nlme)){install.packages("nlme")}
  library("nlme")
if (!requireNamespace("dplyr", quietly = TRUE))  BiocManager::install("dplyr")
library("dplyr")
print("Libraries are loaded.")
```

The metadata contains the processed vs unprocessed info and the order component.

```{r Load and organize metadata}
#### Read metadata ####
# taken from https://app.box.com/file/821046049979
metadata <- openxlsx::read.xlsx(xlsxFile = file.path("data", "NIDDK_linking_LEO_CLEANED_6-11-21.xlsx"),
                                                    sheet = 1)

print(paste("original metadata nrow/ncol:", paste(dim(metadata), collapse = "/")))
metadata <- metadata[metadata$Sample_matrix == "24h Urine",]
print(paste("24h urine metadata nrow/ncol:", paste(dim(metadata), collapse = "/")))
metadata <- na.omit(metadata)
print(paste("metadata no NA nrow/ncol:", paste(dim(metadata), collapse = "/")))
```

```{r Scaling investigation}
#Taken from https://app.box.com/folder/132711394563
initial_table <- data.table::fread(file = file.path("data", "NIDDK_urine_adl_rslts.csv"),
                                   data.table = FALSE)
print(paste("initial rw/col:", paste(dim(initial_table), collapse = "/")))

#### Reorganize and preprocess table ####
initial_table <- initial_table[initial_table$RSLT_TYPE == "Scaled",]
print(paste("filtered rw/col:", paste(dim(initial_table), collapse = "/")))

#Remove non-metabolomic columns
row.names(initial_table) <- initial_table$CLIENT_IDENTIFIER
initial_table <- initial_table[,7:ncol(initial_table)]
print(paste("rn row rw/col:", paste(dim(initial_table), collapse = "/")))

row_median <- apply(initial_table, 1, median, na.rm=TRUE)
col_median <- apply(initial_table, 2, median, na.rm=TRUE)
df_median <- median(as.matrix(initial_table))

print(median(row_median))
```
It appears that the scaling is so that the median of the row medians is 1.


The metadata starts with 294 rows, but some of those are either full of NA
values or not 24h urine samples. After removing these, there are only 80 rows.

Next read in the metabolomics data.

```{r read and organize metabolomics}
#Taken from https://app.box.com/folder/132711394563
initial_table <- data.table::fread(file = file.path("data", "NIDDK_urine_adl_rslts.csv"),
                                   data.table = FALSE)
print(paste("initial rw/col:", paste(dim(initial_table), collapse = "/")))

#### Reorganize and preprocess table ####
initial_table <- initial_table[initial_table$RSLT_TYPE == "Scaled Imp",]
print(paste("filtered rw/col:", paste(dim(initial_table), collapse = "/")))

print(paste("date filtered rw/col:", paste(dim(initial_table), collapse = "/")))

initial_table <- as.data.frame(initial_table)
#Remove non-metabolomic columns
row.names(initial_table) <- initial_table$CLIENT_IDENTIFIER
initial_table <- initial_table[,7:ncol(initial_table)]
print(paste("rn row rw/col:", paste(dim(initial_table), collapse = "/")))

# row_median <- apply(initial_table, 1, median, na.rm=TRUE)
# col_median <- apply(initial_table, 2, median, na.rm=TRUE)
# df_median <- median(as.matrix(initial_table))
# print(paste("row:", row_median, "col:", col_median, "df:", df_median))

#Remove features with more than 80% zeros
initial_table <- initial_table[,sapply(initial_table, function(x) mean(x == 0) < 0.8)]
print(paste("filtered rw/col:", paste(dim(initial_table), collapse = "/")))

#Exclude metabolites that have a CV>30.0
initial_table <- initial_table[,sapply(initial_table, function(x)  sd(x) / mean(x) < 0.7) ]
print(paste("filtered rw/col:", paste(dim(initial_table), collapse = "/")))


#log2 table
initial_table <- log2(initial_table)
```

Metadata has 80 unique samples whereas metabolomics data has 23. This must be
rectified.

```{r Match metadata and metabolomic data}

meta_intrsct <- intersect(row.names(initial_table), metadata$CLIENT_IDENTIFIER)#intersect metadat/metabolomics

metabolomics <- initial_table[meta_intrsct,]

#order both dataframes the same
my_metadata <- metadata[match(metadata$CLIENT_IDENTIFIER, row.names(metabolomics)),]

print(paste("metabolomics rw/col:", paste(dim(metabolomics), collapse = "/")))
identical(my_metadata$CLIENT_IDENTIFIER, row.names(metabolomics))
```

Now that they are matching and we have a final table, add the linear mixed model. It should replicate Lauren's SAS:

Proc mixed data=UrineKB1;
class  ARM Treatment SubjectID;
model METABOLITE= ARM Treatment;
random SubjectID;
estimate ‘ARM LC/LF’ intercept 1 Arm 1 0;
estimate ‘ARM LF/LC’ intercept 1 ARM 0 1;
estimate ‘LC/LF vs LF/LC’ ARM 1 -1;
ods output Estimates = tem;
quit;


https://www.r-bloggers.com/2017/12/linear-mixed-effect-models-in-r/

``` {r Mixed model}

metabolite <- c()
metadata_col <- c()
pval <- c()

for (i in 1:ncol(metabolomics)){
  metabo = unlist(metabolomics[,i])
  metad = as.factor(unlist(metadata$DIET_ORDER))
  si = unlist(metadata$SUBJECT_ID)
  lc_lf <- as.factor(unlist(metadata$DIET))
  lf_lc <-
  myD = data.frame(metabo, metad, pn)
  myLme = lme(metabo ~ metad + , 
    method= "REML", 
    random = ~1 | si,
    # fixed = 
    na.action = 'na.omit',
    data = myD
    )
  myAnova = anova(myLme)
  myPval = myAnova$"p-value"[2]
  metaboliteNames = c(metaboliteNames, row.names(testingData)[i1])
  metadataCatagories = c(metadataCatagories, row.names(metadata)[i2])
  pvals =  c(pvals, myPval)
}

lmeMetadata = function(testingData, label){
  if(!require(nlme)){install.packages("nlme")}
  library("nlme")
  
  metaboliteNames = c()
  metadataCatagories = c()
  pvals = c()
  
  #vessel to hold the failures for future reference
  failedLme <<- data.frame(#'<<-' gives global scope
    metaboliteName = character(0), 
    metaboliteRowNum = numeric(0), 
    metadataName = character(0),
    metadataRowNum = numeric(0)
    )
  
  for (i1 in 1:nrow(testingData)){
    #print(paste('i1:',i1))
    for (i2 in 1:nrow(metadata)){
      #print(paste('i2:',i2))
      if (row.names(metadata)[i2] != 'plant number'){
        tryCatch(
          {
          metabo = unlist(testingData[i1,])
          metad = as.factor(unlist(metadata[i2,]))
          pn = unlist(metadata['plant number',])
          myD = data.frame(metabo, metad, pn)
          myLme = lme(metabo ~ metad, 
            method= "REML", 
            random = ~1 | pn, 
            na.action = 'na.omit',
            data = myD
            )
          myAnova = anova(myLme)
          myPval = myAnova$"p-value"[2]
          metaboliteNames = c(metaboliteNames, row.names(testingData)[i1])
          metadataCatagories = c(metadataCatagories, row.names(metadata)[i2])
          pvals =  c(pvals, myPval)
          },
          error=function(cond) {
            #metaboliteName, metaboliteRowNum, metadataName, metadataRowNum
            myRow = list(row.names(testingData)[i1], i1, row.names(metadata)[i2], i2)
            names(myRow) = c('metaboliteName', 'metaboliteRowNum', 'metadataName', 'metadataRowNum')
           # print(metabo)
            failedLme <<- rbind(failedLme, data.frame(myRow))
            #print('an error is thrown')
            message(cond)
            # Choose a return value in case of error
            # return(NA)
          },
          warning=function(cond) {
            print('a warning is thrown')
            myRow = list(row.names(testingData)[i1], i1, row.names(metadata)[i2], i2)
            names(myRow) = c('metaboliteName', 'metaboliteRowNum', 'metadataName', 'metadataRowNum')
            #print(names(myRow))
            failedLme = rbind(failedLme, data.frame(myRow))
            #print('failedLme warning')
            message(cond)
            # Choose a return value in case of warning
            #return(NULL)
          }
        )
      }
    }
  }
  
  pvalAdj = p.adjust(pvals, method = "BH")
  
  dFrame <- data.frame(pvals,pvalAdj,metaboliteNames,metadataCatagories)
  dFrame <- dFrame[order(dFrame$pvalAdj),]
  
  write.table(dFrame, file=paste(ds, "MetaVs", label, "Lme.txt", sep = ''), row.names=FALSE, sep="\t")
  
  write.table(failedLme, file=paste(ds, "FailedMetaVs", label, "Lme.txt", sep = ''), row.names=FALSE, sep="\t")
  
  pubDf = reshape(dFrame, idvar = 'metaboliteNames', timevar = 'metadataCatagories', direction="wide", sep="", drop = 'pvals')
  write.table(format(pubDf, digits = 3), file=paste(ds, "MetaVs",label,"ReformatShortLme.txt", sep = ''), row.names=FALSE, sep="\t")
  
  pdf(paste(ds, "HistogramMetaVs", label, ".pdf", sep=""))
  hist(pvals, 
    main = paste('Histogram of lme pvalues for ', label),
    xlab = 'Uncorrected pvalues',
    breaks = 50
  )
  dev.off()
  
  pdf(paste(ds, "MetaVs", label, "Lme.pdf", sep=""))
  par(bty = 'l', 
    mar = c(5, 4, 5, 2) + 0.1)
  for( i in 1:nrow(dFrame)){
  	aTitle <- paste(  dFrame$metaboliteNames[i], "vs",  dFrame$metadataCatagories[i], "\nAdjusted Pvalue=", dFrame$pvalAdj[i])
  
  	plot( as.factor(as.character(metadata[as.character(dFrame$metadataCatagories[i]),])),
  	  unlist(testingData[as.character(dFrame$metaboliteNames[i]),]),
  	  main = aTitle,
  	  xlab = as.character(dFrame$metadataCatagories[i]),
  	  ylab = as.character(dFrame$metaboliteNames[i])
  	  )
  }
  dev.off()
  print('successfull run of lmeMetadata')
}#end function

```






