---
title: "Exploratory analysis of ADL metabolomics"
output: "pdf_document"
author: "Aaron Yerke (aaronyerke@gmail.com)"
editor_options: 
  markdown: 
    wrap: 80
---

Combination of urine and plasma from the ultraprocessed vs min processed order effects.

The purpose of this notebook is to explore the metabolomics from the study
below: BACKGROUND: Crossover studies can induce order effects, especially when
they lack a wash-22 out period.23 OBJECTIVE: We aimed to explore diet order
effects on energy balance and food intake24 between randomized diet order groups
in two inpatient crossover studies originally designed to25 compare within
subject differences in ad libitum energy intake between either minimally26
processed low carbohydrate (LC) versus low fat (LF) diets or macronutrient
matched diets27 composed of mostly minimally processed food (MPF) or
ultra-processed food (UPF).

Starting with the URINE 24h from the

```{r setup}
rm(list = ls()) #clear workspace
knitr::opts_knit$set(root.dir = normalizePath(".."))#Set root
knitr::opts_chunk$set(class.source = "Solarized Light")
```

```{r Load external libraries}
#### Loading libraries and data ####
#read in libraries
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("ggplot2", quietly = TRUE))  BiocManager::install("ggplot2")
library("ggplot2")
if (!requireNamespace("readxl", quietly = TRUE))  BiocManager::install("readxl")
library("readxl")
if (!requireNamespace("data.table", quietly = TRUE))  BiocManager::install("data.table")
library("data.table")
if(!require(nlme)){install.packages("nlme")}
  library("nlme")
if(!require("lme4")){install.packages("lme4")}
  library("lme4")
if(!require("poolr")){install.packages("poolr")}
  library("poolr")

print("Libraries are loaded.")
```
```{r functions}

filter_metabolites <- function(df) {
  #Remove columns that all but 2 NA
  df <- df[,!sapply(df, function(x) length(which(is.na(x))) >= nrow(df)-2)]
  
  #Remove features with more than 80% zeros
  # NOTE: Keep NA as NA
  na_table <- df
  na_table[is.na(na_table)] <- 0
  df <- df[,!sapply(na_table, function(x) mean(x == 0) > 0.8)]
  print(paste("0 rw/col:", paste(dim(df), collapse = "/")))
  
  #Exclude metabolites that have a CV>30.0 OR keep metabolites that are < 30
  df <- df[,!sapply(df, 
                                         function(x) sd(x, na.rm = T) / mean(x, na.rm = T) < 0.3)]
  
  print(paste("CV rw/col:", paste(dim(df), collapse = "/")))
  
  return(names(df))
}

simple_model <- function(metabolomics, metadata) {
  metaboliteNames <- c()
  metadata_col <- c()
  pvals <- c()
  failed_metabolites <-c()

  for (i in 1:ncol(metabolomics)){
    metaboliteName <- colnames(metabolomics)[i]
    myD <- metadata
    myD$metabo <- unlist(metabolomics[,i])
    tryCatch(expr = {
            myLme <- nlme::lme( metabo ~ DIET_ORDER + DIET + DIET_ORDER*DIET,
            random = ~ 1 | SUBJECT_ID,
            data = myD)
            #take DIET_ORDER*DIET back out if nothing improves
            #Pull out pvalue for diet_order variable to run through pooled R
            #Look into failed
            metaboliteNames <- c(metaboliteNames, metaboliteName)
            cs <- as.data.frame(summary(myLme)$tTable)
            print(cs)
            if ( !startsWith(row.names(cs)[2], "DIET_ORDER")){
              print("It isn't the first row.")
              print(paste(row.names(cs), collapse = ", "))
            }
            dietorder_pval <- cs[2, "p-value"]
            pvals <- c(pvals, dietorder_pval)
            # myLme = nlme::gls(metabo ~  DIET_ORDER*TIMEPOINT + DIET*TIMEPOINT,
            # na.action=na.omit, data=myD,
            # correlation=nlme::corSymm(form= ~ as.integer(TIMEPOINT) | SUBJECT_ID),
            # weights=nlme::varIdent(form=~1|TIMEPOINT))
            },
            error=function(cond) {
              failed_metabolites <<- c(failed_metabolites, metaboliteName)
              print(paste("an error is thrown on iteration", i, metaboliteName))
              message(cond)
              # Choose a return value in case of error
              # return(NA)
            },
            warning=function(cond) {
              failed_metabolites <<- c(failed_metabolites, metaboliteName)
              print(paste("a warning is thrown on iteration", i,metaboliteName))
              message(cond)
  
              # Choose a return value in case of warning
              # return(NULL)
            }
          )
  }

  names(pvals) <- metaboliteNames
  
  model1_mean_pval <- mean(pvals)
  adj_pvals <- p.adjust(pvals, method = "BH")
  model1_mean_adj_pval <- mean(adj_pvals)
  sig_metabo <- which(adj_pvals < 0.05)
  
  print(paste("mean dietorder pval for model1:", model1_mean_pval, "\n", sum(pvals < 0.05), "significant out of", length(pvals),"metabolites"))
  
  print(paste("mean dietorder adj_pval for model1:", model1_mean_adj_pval, "\n", sum(adj_pvals < 0.05), "significant out of", length(pvals),"metabolites \n", paste(metaboliteNames[sig_metabo], collapse = ", ")))
  
  return(pvals)
}

pathway_loop <- function(result_pv, 
                         pathway_col = "SUPER_PATHWAY",
                         chem_col = "SAS_NAME",
                         pathway_df,
                         metabolomic_df) {#function to combine pvalues based on pathway using Fisher's method
  # inner function for pathway loop
  sup_path_f <- function(result_pv, pathway_col = "SUPER_PATHWAY", chem_col = "SAS_NAME",
                         pathway_name, pathway_df, metabolomic_df){
    # Note: result_pv is the result pvals with the metabolites as the names
  
    # filtering results data set for those in the super pathway of analysis
    # print(paste(names(result_pv)[1:5], collapse = ", "))
    pathway_metabos <- pathway_df[,chem_col][which(pathway_df[,pathway_col] == pathway_name)]
    # print(paste(names(result_pv)[1:5], collapse = ", "))
    # print(length(pathway_metabos))
    # filter pvals to only use those that are found in super pathway
    pval_subset <- result_pv[which(toupper(names(result_pv)) %in% toupper(pathway_metabos))]
    # filter metabolite data
    meta_small <- metabolomic_df[,names(pval_subset)]
    # create correlation matrix
    meta.cor <- cor(meta_small)
    # combine p-values with Fisher's method
    c <- poolr::fisher(pval_subset, adjust = "empirical", R = meta.cor)
    # extracting p-value from result
    z <- c$p
    # print(paste(pathway_name, "combined pval:", z))
    return(z)
  }
  
  df_upper_case <- function(df) {
    return(
    data.frame(lapply(df, function(v) {
    if (is.character(v)) return(toupper(v))
    else return(v)
    })))
  }
  
  set.seed(5627)
  if(!pathway_col %in% pathway_df){
    simpleError(paste("Pathway col not in pathway df:", pathway_col))
  }
  sup_pathways <- unique(pathway_df[,pathway_col])
  sup_pathways <- sup_pathways[nzchar(sup_pathways) & complete.cases(sup_pathways)]
  sup_pvals <- c()
  sup_paths <- c()
  for (sup_p in 1:length(sup_pathways)){
    tryCatch({
      pw_nm <- sup_pathways[sup_p]
      
      my_pv <- sup_path_f(result_pv = result_pv,
                    pathway_col = pathway_col,
                    pathway_name = pw_nm,
                    chem_col = chem_col,
                    pathway_df = pathway_df,
                    metabolomic_df = metabolomic_df)
      sup_pvals <- c(sup_pvals, my_pv)
      sup_paths <- c(sup_paths, pw_nm)
      },
      error=function(cond){
              print(paste("an error is thrown on iteration", sup_p, pw_nm))
              message(cond)
              # Choose a return value in case of error
              # return(NA)
            }
      # warning=function(cond) {
      #   print(paste("a warning is thrown on iteration", sup_p))
      #   message(cond)
      # 
      #   # Choose a return value in case of warning
      #   # return(NULL)
      # }
    )
  }
  names(sup_pvals) <- sup_paths
  model1_mean_pval <- mean(sup_pvals)
  sig_metabo <- which(sup_pvals < 0.05)

  return(sup_pvals)
}


```
The metadata contains the processed vs unprocessed info and the order component.

To build our LMM, we need to have a table with only the following columns:
SUBJECT_ID, ORDER, DIET, TIMEPOINT.

```{r Load and organize metadata}
#### Read metadata ####
# taken from https://app.box.com/file/821046049979
metadata <- openxlsx::read.xlsx(xlsxFile = file.path("data", "NIDDK_linking_LEO_CLEANED_6-11-21.xlsx"), sheet = 1)
print(paste("original metadata nrow/ncol:", paste(dim(metadata), collapse = "/")))
metadata <- metadata[metadata$STUDY_NAME == "ADL",]
metadata <- metadata[metadata$TIMEPOINT %in% c(2,4),]
print(paste("selected tp and study nrow/ncol:", paste(dim(metadata), collapse = "/")))
metadata <- na.omit(metadata)
metadata$SUBJECT_ID <- as.factor(metadata$SUBJECT_ID)
metadata$DIET <- as.factor(metadata$DIET)
metadata$DIET_ORDER <- as.factor(metadata$DIET_ORDER)
metadata$TIMEPOINT <- ordered(as.factor(metadata$TIMEPOINT))

plasma_metadata <- metadata[metadata$Sample_matrix == "PLASMA",]
urine_metadata <- metadata[metadata$Sample_matrix == "24h Urine",]

row.names(plasma_metadata) <- plasma_metadata$CLIENT_IDENTIFIER
row.names(urine_metadata) <- urine_metadata$CLIENT_IDENTIFIER
```


Next read in the urine metabolomics data.

```{r read and organize urine metabolites}

#Taken from https://app.box.com/folder/132711394563
initial_table <- data.frame(data.table::fread(file = file.path("data", "NIDDK_urine_adl_rslts.csv"),
                                   data.table = FALSE))
print(paste("initial rw/col:", paste(dim(initial_table), collapse = "/")))

#### Reorganize and preprocess table ####
initial_table <- initial_table[initial_table$RSLT_TYPE == "Scaled",]
print(paste("scaled rw/col:", paste(dim(initial_table), collapse = "/")))

#Remove non-metabolomic columns
row.names(initial_table) <- initial_table$CLIENT_IDENTIFIER

#### Remove rows that aren't in metadata
initial_table <- merge(urine_metadata, initial_table, 
                     by = 0, 
                     all.x = T)
initial_table <- type.convert(initial_table, as.is = TRUE)#automatically reset incorrectly classified column types

row.names(initial_table) <- initial_table$Row.names

initial_table <- initial_table[,25:ncol(initial_table)]
print(paste("metabolite only rw/col:", paste(dim(initial_table), collapse = "/")))


my_cols <- filter_metabolites(initial_table)
initial_table <- data.frame(data.table::fread(file = file.path("data", "NIDDK_urine_adl_rslts.csv"),
                                   data.table = FALSE))
initial_table <- initial_table[initial_table$RSLT_TYPE == "Scaled Imp",]
row.names(initial_table) <- initial_table$CLIENT_IDENTIFIER

initial_table <- merge(urine_metadata, initial_table, 
                     by = 0, 
                     all.x = T)
initial_table <- type.convert(initial_table, as.is = TRUE)#automatically reset incorrectly classified column types

row.names(initial_table) <- initial_table$Row.names

initial_table <- initial_table[,my_cols]

#log2 table
urine_metabolomics <- log2(initial_table)
```



```{r read and organize plasma metabolomics}

initial_table <- data.frame(data.table::fread(file = file.path("data", "NIDDK_plasma_adl_rslts.csv"),
                                   data.table = FALSE))
print(paste("initial rw/col:", paste(dim(initial_table), collapse = "/")))

#### Reorganize and preprocess table ####
initial_table <- initial_table[initial_table$RSLT_TYPE == "Scaled",]
print(paste("scaled rw/col:", paste(dim(initial_table), collapse = "/")))

#Remove non-metabolomic columns
row.names(initial_table) <- initial_table$CLIENT_IDENTIFIER

#### Remove rows that aren't in metadata
initial_table <- merge(plasma_metadata, initial_table, 
                     by = 0, 
                     all.x = T)
initial_table <- type.convert(initial_table, as.is = TRUE)#automatically reset incorrectly classified column types

row.names(initial_table) <- initial_table$Row.names

initial_table <- initial_table[,25:ncol(initial_table)]
print(paste("metabolite only rw/col:", paste(dim(initial_table), collapse = "/")))

#### Filter and switch to scaled Imp ####
my_cols <- filter_metabolites(initial_table)

initial_table <- data.frame(data.table::fread(file = file.path("data", "NIDDK_plasma_adl_rslts.csv"),
                                   data.table = FALSE))
initial_table <- initial_table[initial_table$RSLT_TYPE == "Scaled Imp",]
row.names(initial_table) <- initial_table$CLIENT_IDENTIFIER

initial_table <- merge(plasma_metadata, initial_table, 
                     by = 0, 
                     all.x = T)
initial_table <- type.convert(initial_table, as.is = TRUE)#automatically reset incorrectly classified column types

row.names(initial_table) <- initial_table$Row.names

initial_table <- initial_table[,my_cols]

#log2 table
plasma_metabolomics <- log2(initial_table)
```



Now that they are matching and we have a final table, add the linear mixed
model. It should replicate Lauren's SAS:

Proc mixed data=UrineKB1; class ARM Treatment SubjectID; model METABOLITE= ARM
Treatment; random SubjectID; estimate 'ARM LC/LF' intercept 1 Arm 1 0; estimate
'ARM LF/LC' intercept 1 ARM 0 1; estimate 'LC/LF vs LF/LC' ARM 1 -1; ods output
Estimates = tem; quit;

<https://www.r-bloggers.com/2017/12/linear-mixed-effect-models-in-r/>
<https://mspeekenbrink.github.io/sdam-r-companion/linear-mixed-effects-models.html>

```{r Mixed model}
#### A little more data organisation ####
plasma_metadata <- plasma_metadata[, c("SUBJECT_ID", "DIET", "DIET_ORDER")]
print(paste("final plasma_metadata r/c:", paste(dim(plasma_metadata), collapse = "/")))

row.names(urine_metadata) <- urine_metadata$CLIENT_IDENTIFIER
urine_metadata <- urine_metadata[, c("SUBJECT_ID", "DIET", "DIET_ORDER")]
print(paste("final urine_metadata r/c:", paste(dim(urine_metadata), collapse = "/")))

# run the model
urine_pvals <- simple_model(urine_metabolomics, urine_metadata)
plasma_pvals <- simple_model(plasma_metabolomics, plasma_metadata)
```

Add the pvalue pooling from Erikka's code.

```{r pathway analyses}

#Load pathway data
urine_metabo_path <- data.table::fread(file = file.path("data", "NIDDK_urine_metabolites.csv"),
                                   data.table = FALSE)
plasma_metabo_path <- data.table::fread(file = file.path("data", "NIDDK_plasma_metabolites.csv"),
                                   data.table = FALSE)

# Super pathways
adl_urine_sup_pv <- pathway_loop(result_pv = urine_pvals,
                             pathway_col = "SUPER_PATHWAY",
                             pathway_df = urine_metabo_path,
                             metabolomic_df = urine_metabolomics)
adl_plasma_sup_pv <- pathway_loop(result_pv = plasma_pvals,
                             pathway_col = "SUPER_PATHWAY",
                             pathway_df = plasma_metabo_path,
                             metabolomic_df = plasma_metabolomics)   
    
# Sub pathways
adl_urine_sub_pv <- pathway_loop(result_pv = urine_pvals,
                             pathway_col = "SUB_PATHWAY",
                             pathway_df = urine_metabo_path,
                             metabolomic_df = urine_metabolomics)
adl_plasma_sub_pv <- pathway_loop(result_pv = plasma_pvals,
                             pathway_col = "SUB_PATHWAY",
                             pathway_df = plasma_metabo_path,
                             metabolomic_df = plasma_metabolomics)   
```

Organization of the LCLF data:
The metabolomics data links to the metadata through the 

	Plasma metadata	plasma metabolomics data	plasma metabolomics data
FILE	SampleLabIDs.xlsx	NCIA-0106-20PHML+ DATA TABLES (PLASMA)_CURRENT.xlsx	NCIA-0106-20PHML+ DATA TABLES (PLASMA)_CURRENT.xlsx
SHEET	NA	Sample Meta Data	Batch-normalized Data
COLUMN	CLIENT_SAMPLE_ID	CLIENT_SAMPLE_ID	PARENT_SAMPLE_NAME
DATA EX	LCF000248 0002	LCF000038 0002	NCIA-34649


```{r Load and organize lclf plasma metadata}
#### Read metadata ####
#metabolon metabolomics file that contains most of the data in several sheets
metabolon_file <- file.path("data", "NCIA-0106-20PHML+", "NCIA-0106-20PHML+ (PLASMA) MERGED.xlsx")

lclf_plas_metad <- openxlsx::read.xlsx(xlsxFile = file.path("data", "SampleLabIDs.xlsx"),
                                     sheet = 1)
print(paste("original lclf_plas_metad nrow/ncol:", paste(dim(lclf_plas_metad), collapse = "/")))

lclf_samp_link <- openxlsx::read.xlsx(xlsxFile = metabolon_file,
                                      sheet = "Merged Sample Meta Data")
print(paste("original lclf_samp_link nrow/ncol:", paste(dim(lclf_samp_link), collapse = "/")))

lclf_plas_metad <- na.omit(lclf_plas_metad)

print(paste("na lclf_plas_metad nrow/ncol:", paste(dim(lclf_plas_metad), collapse = "/")))

lclf_plas_metad$PARENT_SAMPLE_NAME <- lclf_samp_link$PARENT_SAMPLE_NAME[match(lclf_plas_metad$CLIENT_SAMPLE_ID, lclf_samp_link$CLIENT_SAMPLE_ID)]

lclf_plas_metad <- na.omit(lclf_plas_metad, cols = "PARENT_SAMPLE_NAME")

row.names(lclf_plas_metad) <- lclf_plas_metad$PARENT_SAMPLE_NAME

lclf_plas_metad$SUBJECT_ID <- as.factor(lclf_plas_metad$SubjectID)
lclf_plas_metad$DIET <- as.factor(lclf_plas_metad$Diet)
lclf_plas_metad$DIET_ORDER <- as.factor(lclf_plas_metad$ARM)

lclf_plas_metad <- lclf_plas_metad[, c("SUBJECT_ID", "DIET", "DIET_ORDER")]

print(paste("final lclf_plas_metad nrow/ncol:", paste(dim(lclf_plas_metad), collapse = "/")))
```


```{r Load and organize lclf metabolomics}
#### Read lclf data ####
lclf_plasma_mtbmcs <- openxlsx::read.xlsx(xlsxFile = metabolon_file,
                                     sheet = "Merged Batch Norm NonImp",
                                     rowNames = TRUE)
print(paste("original lclf_plasma_mtbmcs nrow/ncol:", paste(dim(lclf_plasma_mtbmcs), collapse = "/")))

lclf_chem_link <- openxlsx::read.xlsx(xlsxFile = metabolon_file,
                                      sheet = "Merged Annotation")

names(lclf_plasma_mtbmcs) <- lclf_chem_link$CHEMICAL_NAME[match(names(lclf_plasma_mtbmcs), lclf_chem_link$CHEM_ID)]

lclf_plasma_mtbmcs <- merge(lclf_plas_metad, lclf_plasma_mtbmcs, by = 0)
row.names(lclf_plasma_mtbmcs) <- lclf_plasma_mtbmcs$Row.names
lclf_plasma_mtbmcs <- lclf_plasma_mtbmcs[,(ncol(lclf_plas_metad)+2):ncol(lclf_plasma_mtbmcs)]

print(paste("final lclf_plasma_mtbmcs nrow/ncol:", paste(dim(lclf_plasma_mtbmcs), collapse = "/")))

my_cols <- filter_metabolites(lclf_plasma_mtbmcs)

lclf_plasma_mtbmcs <- openxlsx::read.xlsx(xlsxFile = metabolon_file, 
                                          sheet = "Merged Batch Norm Imp",
                                          rowNames = TRUE)
names(lclf_plasma_mtbmcs) <- lclf_chem_link$CHEMICAL_NAME[match(names(lclf_plasma_mtbmcs), lclf_chem_link$CHEM_ID)]
lclf_plasma_mtbmcs <- merge(lclf_plas_metad, lclf_plasma_mtbmcs, by = 0)
row.names(lclf_plasma_mtbmcs) <- lclf_plasma_mtbmcs$Row.names
lclf_plasma_mtbmcs <- lclf_plasma_mtbmcs[,my_cols]

```


```{r model and pvals}
lclf_plasma_pvals <- simple_model(lclf_plasma_mtbmcs, lclf_plas_metad)

lclf_plasma_sup_pv <- pathway_loop(result_pv = lclf_plasma_pvals,
                             pathway_col = "SUPER_PATHWAY",
                             chem_col = "CHEMICAL_NAME",
                             pathway_df = lclf_chem_link,
                             metabolomic_df = lclf_plasma_mtbmcs)

# Sub pathways
lclf_plasma_sub_pv <- pathway_loop(result_pv = lclf_plasma_pvals,
                             pathway_col = "SUB_PATHWAY",
                             chem_col = "CHEMICAL_NAME",
                             pathway_df = lclf_chem_link,
                             metabolomic_df = lclf_plasma_mtbmcs)
```


```{r Load and organize lclf urine metadata}
#### Read metadata ####
#metabolon metabolomics file that contains most of the data in several sheets
metabolon_file <- file.path("data", "NCIA-0106-20PHML+", "NCIA-0106-20PHML+ (URINE) MERGED.xlsx")
metadata_file <- file.path("data", "NCIA-0106-20PHML+", "SampleLabIDs20231101.xlsx")

lclf_urin_metad<- openxlsx::read.xlsx(xlsxFile = metadata_file,
                                     sheet = 1)
print(paste("original lclf_urin_metadnrow/ncol:", paste(dim(lclf_urin_metad), collapse = "/")))

lclf_samp_link <- openxlsx::read.xlsx(xlsxFile = metabolon_file,
                                      sheet = "Merged Sample Meta Data")
print(paste("original lclf_samp_link nrow/ncol:", paste(dim(lclf_samp_link), collapse = "/")))

lclf_urin_metad <- lclf_urin_metad[lclf_urin_metad$Material == "24h Urine",]

lclf_urin_metad<- na.omit(lclf_urin_metad)

print(paste("na lclf_urin_metadnrow/ncol:", paste(dim(lclf_urin_metad), collapse = "/")))

lclf_urin_metad$PARENT_SAMPLE_NAME <- lclf_samp_link$PARENT_SAMPLE_NAME[match(lclf_urin_metad$CLIENT_SAMPLE_ID, lclf_samp_link$CLIENT_SAMPLE_ID)]

lclf_urin_metad<- na.omit(lclf_urin_metad, cols = "PARENT_SAMPLE_NAME")

row.names(lclf_urin_metad) <- lclf_urin_metad$PARENT_SAMPLE_NAME

lclf_urin_metad$SUBJECT_ID <- as.factor(lclf_urin_metad$SubjectID)
lclf_urin_metad$DIET <- as.factor(lclf_urin_metad$Diet)
lclf_urin_metad$DIET_ORDER <- as.factor(lclf_urin_metad$ARM)

lclf_urin_metad<- lclf_urin_metad[, c("SUBJECT_ID", "DIET", "DIET_ORDER")]

print(paste("final lclf_urin_metadnrow/ncol:", paste(dim(lclf_urin_metad), collapse = "/")))
```

```{r Load and organize lclf urine metabolomics}
#### Read lclf data ####
lclf_urin_mtbmcs <- openxlsx::read.xlsx(xlsxFile = metabolon_file,
                                     sheet = "Merged Batch Norm NonImp",
                                     rowNames = TRUE)
print(paste("original lclf_urin_mtbmcs nrow/ncol:", paste(dim(lclf_urin_mtbmcs), collapse = "/")))

lclf_chem_link <- openxlsx::read.xlsx(xlsxFile = metabolon_file,
                                      sheet = "Merged Annotation")

names(lclf_urin_mtbmcs) <- lclf_chem_link$CHEMICAL_NAME[match(names(lclf_urin_mtbmcs), lclf_chem_link$CHEM_ID)]

lclf_urin_mtbmcs <- merge(lclf_urin_metad, lclf_urin_mtbmcs, by = 0)
row.names(lclf_urin_mtbmcs) <- lclf_urin_mtbmcs$Row.names
lclf_urin_mtbmcs <- lclf_urin_mtbmcs[,(ncol(lclf_urin_metad)+2):ncol(lclf_urin_mtbmcs)]

print(paste("final lclf_urin_mtbmcs nrow/ncol:", paste(dim(lclf_urin_mtbmcs), collapse = "/")))

my_cols <- filter_metabolites(lclf_urin_mtbmcs)

lclf_urin_mtbmcs <- openxlsx::read.xlsx(xlsxFile = metabolon_file, 
                                          sheet = "Merged Batch Norm Imp",
                                          rowNames = TRUE)
names(lclf_urin_mtbmcs) <- lclf_chem_link$CHEMICAL_NAME[match(names(lclf_urin_mtbmcs), lclf_chem_link$CHEM_ID)]
lclf_urin_mtbmcs <- merge(lclf_urin_metad, lclf_urin_mtbmcs, by = 0)
row.names(lclf_urin_mtbmcs) <- lclf_urin_mtbmcs$Row.names
lclf_urin_mtbmcs <- lclf_urin_mtbmcs[,my_cols]

```


```{r model and pvals}

lclf_urin_pvals <- simple_model(lclf_urin_mtbmcs, lclf_urin_metad)

lclf_urin_sup_pv <- pathway_loop(result_pv = lclf_urin_pvals,
                             pathway_col = "SUPER_PATHWAY",
                             chem_col = "CHEMICAL_NAME",
                             pathway_df = lclf_chem_link,
                             metabolomic_df = lclf_urin_mtbmcs)

# Sub pathways
lclf_urin_sub_pv <- pathway_loop(result_pv = lclf_urin_pvals,
                             pathway_col = "SUB_PATHWAY",
                             chem_col = "CHEMICAL_NAME",
                             pathway_df = lclf_chem_link,
                             metabolomic_df = lclf_urin_mtbmcs)


```

```{r figures}
library("ggplot2")

my_df <- data.frame("pathways" = c(names(adl_plasma_sup_pv),
                                   names(adl_urine_sup_pv),
                                   names(lclf_plasma_sup_pv),
                                   names(lclf_urin_sup_pv)), 
                    "pval" = -log10(c(adl_plasma_sup_pv,
                                      adl_urine_sup_pv,
                                      lclf_plasma_sup_pv,
                                      lclf_urin_sup_pv)),
                    "tissue" = c(rep("ADL plasma", length(adl_plasma_sup_pv)), 
                                 rep("ADL urine", length(adl_urine_sup_pv)), 
                                 rep("LCLF plasma", length(lclf_plasma_sup_pv)),
                                 rep("LCLF urine", length(lclf_urin_sup_pv))))

axis_font_size <- 12
font_type <- "TT Arial"

ggplot(my_df, aes(x = pathways, y = pval, color = factor(tissue))) +
  ggplot2::geom_hline(yintercept = -log10(0.05), color = "red") +
  geom_point(aes(shape = factor(tissue), color = factor(tissue))) +
  labs(colour = "Legend") +
  ggtitle(paste0("ADL Super pathway analysis")) + # Blank Title for the Graph
  xlab("Super pathways") +
  ylab(paste("-log(pvalue)")) +
  theme(line = element_line(colour = "black", linewidth = 1), 
        # strip.background = element_rect(fill="gray90"),
        axis.text.x = element_text(angle = 90, hjust=1),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        # panel.border= element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  labs(color  = "Legend", shape = "Legend")
```


```{r sub figures}
library("ggplot2")

my_df <- data.frame("pathways" = c(names(adl_plasma_sub_pv), 
                                   names(adl_urine_sub_pv),
                                   names(lclf_plasma_sub_pv),
                                   names(lclf_urin_sub_pv)), 
                    "pval" = -log10(c(adl_plasma_sub_pv,
                                      adl_urine_sub_pv,
                                      lclf_plasma_sub_pv,
                                      lclf_urin_sub_pv)),
                    "tissue" = c(rep("ADL plasma", length(adl_plasma_sub_pv)),
                                 rep("ADL urine", length(adl_urine_sub_pv)), 
                                 rep("LCLF plasma", length(lclf_plasma_sub_pv)),
                                 rep("LCLF urine", length(lclf_urin_sub_pv))))

axis_font_size <- 12
font_type <- "TT Arial"

g <- ggplot(my_df, aes(x = pathways, y = pval, color = factor(tissue))) +
  ggplot2::geom_hline(yintercept = -log10(0.05), color = "red") +
  geom_point(aes(shape = factor(tissue), color = factor(tissue))) +
  labs(colour = "Legend") +
  ggtitle(paste0("ADL sub pathway analysis")) + # Blank Title for the Graph
  xlab("Sub pathways") +
  ylab(paste("-log(pvalue)")) +
  theme(line = element_line(colour = "black", linewidth = 1), 
        # strip.background = element_rect(fill="gray90"),
        axis.text.x = element_text(angle = 90, hjust=1, size = 6),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        # panel.border= element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  labs(color  = "Legend", shape = "Legend")
  # geom_rect(aes(xmin=pval > -log10(0.05), xmax=pval < -log10(0.05), 
  #           ymin=1,ymax=max),
  #               color="transparent", fill="orange", alpha=0.3)

ggplot2::ggsave(plot = g, filename = "sub_path.png",   width = 12, height = 4)

g

#Color by dataset
#Symbol by tissue type
#group x axis by super pathways - might help with interpretation
#try to add rectangles


```



